Male-biased sexual selection, but not sexual dichromatism, predicts speciation

### 2019-08-11 Manuscript_Publication.Rmd
addition of polygyny - speciation rate reference

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§}
  and Iliana Medina\textsuperscript{§}
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  word_document: default
  pdf_document: default
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang,
  flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{colorlinks=true, hyperfootnotes=false,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/NComms.csl
subparagraph: yes
abstract: Sexual selection is thought to shape phylogenetic diversity by affecting
  speciation or extinction rates. However, the net effect of sexual selection on diversification
  is hard to predict _a priori_, because many of the hypothesised effects on speciation
  or extinction have opposing signs and uncertain magnitudes. Here, we test whether
  the strength of sexual selection predicts variation in speciation and extinction
  rates across passerine birds (up to 5,812 species, covering most genera). We tested
  for associations between speciation or extinction and two measures of sexual selection
  (sexual dichromatism and a multivariate measure of male-biased sexual selection),
  accounting for range size and measures of environmental variability. Male-biased
  sexual selection, but not sexual dichromatism, predicted speciation rates ($\lambda_{DR}$)
  in passerines. This relationship was independent of range size or environmental
  variability, though species with smaller ranges had higher speciation rates. There
  was no association between sexual selection and extinction rate ($\mu_{BAMM}$).
  Our findings show that sexual dichromatism is a poor proxy for sexual selection
  at least a broad-scale and support the view that sexual selection has shaped diversification
  in songbirds.
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia}

@@

Sexual selection is a fundamental evolutionary process; yet there is long-standing debate about how it shapes patterns of species diversity[@Lande_1981; @Lande_1982; @west1983sexual; @Seddon_2008; @Cooney_2018]. Sexual selection can promote speciation because it operates on traits that can create reproductive isolation when they diverge between lineages, such as signals and preferences involved in mate selection[@Lande_1981; @Lande_1982; @Safran_2013], sperm-egg interactions [@swanson1998abalone], or genital morphology [@sloan2019genitalia]. Sexual selection could also promote speciation or prevent extinction by purging deleterious mutations[@Whitlock_Agrawal_2009], fixing beneficial ones[@Whitlock_2000], and accelerating adaptation in different environments[@Lorch_2003; @Candolin_2008; @Cally_2019]. Conversely, sexual selection might hinder speciation or make extinction more likely by favouring traits that improve mating success but reduce population fitness[@Rankin_2011; @Kokko_Jennions2008; @Holman_Kokko_2013; @Fromhage_2016]. For example, species with costly sexual signals may be less resilient to environmental change[@Kokko_Brooks_2003]. Sexual selection might also promote extinction by causing maladaptation (‘gender load’) in female traits that are genetically correlated with sexually-selected male traits[@Pischedda_2006; @harano2010intralocus; @Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013]. Although numerous studies have examined the relationship between sexual selection and speciation or extinction rates[@Barraclough_1995; @Morrow_Speciation_2003; @Seddon_2008; @Kraaijeveld_2011; @Seddon_2013; @Huang_2014], the availability of more complete phenotypic, ecological and phylogenetic data[@Jetz_2012], together with significant advances in phylogenetic methods[@Rabosky_2014; @Harvey_essim_2017], present new opportunities to test whether and how sexual selection drives diversification.

@@

Birds have been a popular focus of macroevolutionary studies of sexual selection and diversification [@Barraclough_1995; @Morrow_Speciation_2003; @Seddon_2008; @Seddon_2013; @Huang_2014] because their biology and phylogenetic relationships are comparatively well-known. A 2011 meta-analysis covering 20 primary studies of birds and other taxa found a small but significant positive association between sexual selection and speciation, with the average effect size in birds stronger than in mammals but weaker than in insects or fish [@Kraaijeveld_2011]. However, there was large variation in effect size estimates across the 20 studies, likely reflecting differences in methodology, such as metrics used to characterise speciation and sexual selection, in addition to true biological differences. More recently, Huang & Rabosky [@Huang_2014] found no association between sexual dichromatism and speciation (*n* = 918 species) in a study using spectrophotometric measurements of museum specimens [@Armenta_2008] and tip-rate estimates from a molecular-only phylogeny[@Jetz_2012]. Similarly, Cooney et al. [@Cooney_2017] found no effect of sexual dichromatism on diversification across 1,306 pairs of species, using dichromatism scores provided by human observers.

@@

We examined the effect of sexual selection on speciation and extinction rate in 97% of passerines (*n* = 5,812 species; 58% of all birds). We calculated three different tip-rate metrics of speciation and one of extinction for a maximum clade credibility (MCC) tree and, to account for phylogenetic uncertainty, 1,000 trees for $\lambda_{DR}$ and $\lambda_{ND}$ and 100 trees for $\lambda_{BAMM}$ [@Rabosky_2014]. $\lambda_{DR}$ is a measure of speciation rate more heavily weighted to recent speciation events while $\lambda_{ND}$ measures speciation across the root-to-tip path. Alternatively, $\lambda_{BAMM}$ uses a Bayesian approach to assess the probability of evolutionary rate-shift configurations, from which it generates tip-rate speciation and extinction estimates.

>>>>

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§}
  and Iliana Medina\textsuperscript{§}
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  pdf_document: default
  word_document: default
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang,
  flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{colorlinks=true, hyperfootnotes=false,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/NComms.csl
subparagraph: yes
abstract: Sexual selection is thought to shape phylogenetic diversity by affecting
  speciation or extinction rates. However, the net effect of sexual selection on diversification
  is hard to predict _a priori_, because many of the hypothesised effects on speciation
  or extinction have opposing signs and uncertain magnitudes. Here, we test whether
  the strength of sexual selection predicts variation in speciation and extinction
  rates across passerine birds (up to 5,812 species, covering most genera). We tested
  for associations between speciation or extinction and two measures of sexual selection
  (sexual dichromatism and a multivariate measure of male-biased sexual selection),
  accounting for range size and measures of environmental variability. Male-biased
  sexual selection, but not sexual dichromatism, predicted speciation rates ($\lambda_{DR}$)
  in passerines. This relationship was independent of range size or environmental
  variability, though species with smaller ranges had higher speciation rates. There
  was no association between sexual selection and extinction rate ($\mu_{BAMM}$).
  Our findings show that sexual dichromatism is a poor proxy for sexual selection
  at least a broad-scale and support the view that sexual selection has shaped diversification
  in songbirds.
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia}

@@

Sexual selection is a fundamental evolutionary process; yet there is long-standing debate about how it shapes patterns of species diversity[@Lande_1981; @Lande_1982; @west1983sexual; @Seddon_2008; @Cooney_2018]. Sexual selection can promote speciation because it operates on traits that can create reproductive isolation when they diverge between lineages, such as signals and preferences involved in mate selection[@Lande_1981; @Lande_1982; @Safran_2013], sperm-egg interactions [@swanson1998abalone], or genital morphology [@sloan2019genitalia]. Sexual selection could also promote speciation or prevent extinction by purging deleterious mutations[@Whitlock_Agrawal_2009], fixing beneficial ones[@Whitlock_2000], and accelerating adaptation in different environments[@Lorch_2003; @Candolin_2008; @Cally_2019]. Conversely, sexual selection might hinder speciation or make extinction more likely by favouring traits that improve mating success but reduce population fitness[@Rankin_2011; @Kokko_Jennions2008; @Holman_Kokko_2013; @Fromhage_2016]. For example, species with costly sexual signals may be less resilient to environmental change[@Kokko_Brooks_2003]. Extinction risk may also be exacerbated by sexual selection causing maladaptation (‘gender load’) in female traits that are genetically correlated with sexually-selected male traits [@Pischedda_2006; @harano2010intralocus; @Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013]. Although numerous studies have examined the relationship between sexual selection and speciation or extinction rates [@Barraclough_1995; @Morrow_Speciation_2003; @Seddon_2008; @Kraaijeveld_2011; @Seddon_2013; @Huang_2014], the availability of more complete phenotypic, ecological and phylogenetic data[@Jetz_2012], together with significant advances in phylogenetic methods[@Rabosky_2014; @Harvey_essim_2017], present new opportunities to test whether and how sexual selection drives diversification.

@@

Birds have been a popular focus of macroevolutionary studies of sexual selection and diversification [@Barraclough_1995; @Morrow_Speciation_2003; @Seddon_2008; @Seddon_2013; @Huang_2014] because their biology and phylogenetic relationships are comparatively well-known. A 2011 meta-analysis covering 20 primary studies of birds and other taxa found a small but significant positive association between sexual selection and speciation, with the average effect size in birds stronger than in mammals but weaker than in insects or fish [@Kraaijeveld_2011]. However, there was large variation in effect size estimates across the 20 studies, likely reflecting differences in methodology, such as metrics used to characterise speciation and sexual selection, in addition to true biological differences. More recently, Huang & Rabosky [@Huang_2014] found no association between sexual dichromatism and speciation (*n* = 918 species) in a study using spectrophotometric measurements of museum specimens [@Armenta_2008] and tip-rate estimates from a molecular-only phylogeny[@Jetz_2012]. Similarly, Cooney et al. [@Cooney_2017] found no effect of sexual dichromatism on diversification across 1,306 pairs of species, using dichromatism scores provided by human observers. However, another study using the prevelance of polygyny (as oppossed to sexual dichromatism) as a proxy of sexual selection found a positive association with speciation rate across 954 species of birds [@Iglesias-Carrasco_2019].

@@

We examined the effect of sexual selection on speciation and extinction rate in 97% of passerines (*n* = 5,812 species; 58% of all birds). We calculated three different tip-rate metrics of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$) and one of extinction ($\mu_{BAMM}$) for a maximum clade credibility (MCC) tree and, to account for phylogenetic uncertainty, 1,000 trees for $\lambda_{DR}$ and $\lambda_{ND}$ and 100 trees for $\lambda_{BAMM}$ [@Rabosky_2014]. $\lambda_{DR}$ is a measure of speciation rate more heavily weighted to recent speciation events while $\lambda_{ND}$ measures speciation across the root-to-tip path. Alternatively, $\lambda_{BAMM}$ uses a Bayesian approach to assess the probability of evolutionary rate-shift configurations, from which it generates tip-rate speciation and extinction estimates.
